---
title: Toxicity
author: Jessica Weiss
output:
  html_document:
    mathjax:  default
    fig_caption:  true
    toc: true
    section_numbering: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Creating swimmers plots with ease}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(swimplot)
```

## Sample data

 A sample dataset displaying three types of toxicity for 5 patients over the course of a study
```{r, echo = TRUE,fig.align='centre'}
d1 <- data.frame(pat=c('a','b','c','d','e'),end=c(3.26,15.45,14.84,3.51,7.44))

Nausea <- data.frame(pat=c('a','b','c','c','e'),tox="Nausea",grade=as.character(c(1,3,2,2,3)),time=c(2,1,8,10,1.5))
Fever <- data.frame(pat=c('d','e','e','c','c'),tox="Fever",grade=as.character(c(1,2,2,1,2)), time=c(1.5,3,6,2,6.5))
Anemia <- data.frame(pat=c('d','e','e','c','c'),tox="Anemia",grade=as.character(c(1,1,2,1,3)), time=c(1.5,3,5,2,6))

```

Below are three different examples of ways to represent this data in a swimmers plot

## Example 1

Using numbers to represent the grade of the toxicity, and colour for the type
```{r, echo = TRUE,fig.align='centre'}

swimmer_plot(df=d1,id='pat',end='end',col="black",fill='white')+
  swimmer_points(df_points=Nausea,id='pat',time='time',adj.y = 0.3,size=4,name_shape = 'grade',name_col = 'tox',stroke=3)+
  swimmer_points(df_points=Fever,id='pat',time='time',adj.y = 0,size=4,name_shape = 'grade',name_col = 'tox',stroke=3)+
  swimmer_points(df_points=Anemia,id='pat',time='time',adj.y = -0.3,size=4,name_shape = 'grade',name_col = 'tox',stroke=3)+
  ggplot2::scale_shape_manual(values=c("1","2","3")) + ggplot2::guides(shape=FALSE)+
  ggplot2::scale_color_brewer(palette = "Dark2")
```

## Example 2

Using colour to represent the grade of the toxicity, and shape for the type
```{r, echo = TRUE,fig.align='centre'}

swimmer_plot(df=d1,id='pat',end='end',col="black",fill='white')+
  swimmer_points(df_points=Nausea,id='pat',time='time',adj.y = 0.3,size=4,name_col = 'grade',name_shape = 'tox',stroke=3)+
  swimmer_points(df_points=Fever,id='pat',time='time',adj.y = 0,size=4,name_col = 'grade',name_shape = 'tox',stroke=3)+
  swimmer_points(df_points=Anemia,id='pat',time='time',adj.y = -0.3,size=4,name_col = 'grade',name_shape = 'tox',stroke=3)+ 
  ggplot2::scale_color_manual(name="Grade",values=c("coral2","red3","orangered4"))
```

## Example 3

Using colour and shape to represent the grade of the toxicity, and adding text to represent the type
```{r, echo = TRUE,fig.align='centre'}
textdf1 <- data.frame(pat=unique(d1$pat),label="Nausea")
textdf2 <- data.frame(pat=unique(d1$pat),label="Fever")
textdf3 <- data.frame(pat=unique(d1$pat),label="Anemia")

swimmer_plot(df=d1,id='pat',end='end',col="black",fill='white')+
  swimmer_points(df_points=Nausea,id='pat',time='time',adj.y = 0.2,size=4,name_col = 'grade',name_shape = 'grade',stroke=3)+
  swimmer_points(df_points=Fever,id='pat',time='time',adj.y = 0,size=4,name_col = 'grade',name_shape = 'grade',stroke=3)+
  swimmer_points(df_points=Anemia,id='pat',time='time',adj.y = -0.2,size=4,name_col = 'grade',name_shape = 'grade',stroke=3)+ 
  ggplot2::scale_color_manual(name="Grade",values=c("coral2","red3","orangered4")) + ggplot2::labs(shape="Grade")+
  swimmer_text(df_text = textdf1,id='pat',label = 'label',adj.x = -2.2,adj.y = 0.2)+
  swimmer_text(df_text = textdf2,id='pat',label = 'label',adj.x = -2.2,adj.y = 0)+
  swimmer_text(df_text = textdf3,id='pat',label = 'label',adj.x = -2.2,adj.y = -0.2)
```

